Aligning marine species range data to better serve science and conservation

### 2016-03-04 ms_1_abst_sig.Rmd

Species distribution data provide the foundation for a wide range of ecological research and conservation science and management, yet most species ranges remain unknown, and existing range maps have little overlap. In the ocean, two global species distribution datasets, produced by AquaMaps (REF) and the International Union for Conservation of Nature (IUCN) (REF), dominate our understanding of marine species ranges throughout the world's oceans.  Together they represent `r format(nrow(spp_all), big.mark = ',')` species, mostly from AquaMaps, with only `r format(nrow(spp_map_pairs), big.mark = ',')` species overlapping.  Here we examine differences in predicted species ranges between the two datasets, propose mechanistic causes and potential solutions for such differences, and explore the implications of these differences for management and conservation decisions.  We find that IUCN maps often disregard bathymetry for depth-limited species, leading to predictions of species presence at unsuitable depths, and AquaMaps ranges for data-poor species often extrapolate presence far afield from known occurrences.  We illustrate the implications of these differences by repeating two recent applications - the Ocean Health Index (Halpern et al. 2012, 2015) biodiversity goal and a global analysis of gaps in coverage of marine protected areas (Klein et al. 2015) - and find significantly different estimates of the status of global biodiversity and effectiveness of conservation depending on how the two datasets were prioritized. ___Creating a single, highly accurate dataset of global marine species ranges will be difficult. Understanding the implications of dataset differences for conservation planning and decision-making is essential.___
<!--Understanding these issues points toward solutions for providers and users of these datasets.-->

>>>>

Species distribution data provide the foundation for a wide range of ecological research and conservation science and management, yet most species ranges remain unknown, and existing range maps have little overlap. In the ocean, two global species distribution datasets, produced by AquaMaps (REF) and the International Union for Conservation of Nature (IUCN) (REF), dominate our understanding of marine species ranges throughout the world's oceans.  Together they represent `r format(n_spp_all, big.mark = ',')` species, mostly from AquaMaps, with only `r format(n_spp_both, big.mark = ',')` species overlapping.  Here we examine differences in predicted species ranges between the two datasets, propose mechanistic causes and potential solutions for such differences, and explore the implications of these differences for management and conservation decisions.  We find that IUCN maps often disregard bathymetry for depth-limited species, leading to predictions of species presence at unsuitable depths, and AquaMaps ranges for data-poor species often extrapolate presence far afield from known occurrences.  We illustrate the implications of these differences by repeating two recent applications - the Ocean Health Index (Halpern et al. 2012, 2015) biodiversity goal and a global analysis of gaps in coverage of marine protected areas (Klein et al. 2015) - and find significantly different estimates of the status of global biodiversity and effectiveness of conservation depending on how the two datasets were prioritized. ___Creating a single, highly accurate dataset of global marine species ranges will be difficult. Understanding the implications of dataset differences for conservation planning and decision-making is essential.___

<!--Understanding these issues points toward solutions for providers and users of these datasets.-->

### 2016-03-04 ms_3_methods.Rmd

**Methods for MPA Gap Analysis case study:** 
Based upon the methods described in Klein et al. 2015, we reconstructed the analysis using the subset of protected areas spatially covering a marine area and classified as IUCN I-IV ___(Klein paper, table 1, subset 4 - seems to be the one they focus on the most)___, converting the World Database of Protected Areas (WDPA, REF) dataset to a raster of 0.5° cells indicating proportion of protected area within the cell ___(consider also: filter on STATUS == 'Designated', and filter out those pesky fishery management areas and manatee speed zones...)___.  To verify our reconstruction we ran the analysis using the AquaMaps dataset available in April 2015 (REF) at a presence threshold of 50%.  We then updated the analysis using the most recent AquaMaps dataset (REF?) at 50% threshold

* Possible analyses:
    * run AquaMaps (new data) with threshold at 0%
        * already done in original paper for old data, but redo it to compare to AM + IUCN scenario below
    * run with AquaMaps only, but filtering out data-poor species?
    * run with IUCN maps only in place of AquaMaps?
    * run with AquaMaps and IUCN both, using 0% threshold; which preferred? 
        * Should we clip corals (and coral-associated taxa) to 200 meter bathymetry, just for fun?
* Which aspects of the analysis should we focus on?:
    * overall vs broken down by taxonomic groups? (see paper, figure 1)
    * spatial distribution of gap species (see paper, figure 2)

>>>>

**Methods for MPA Gap Analysis case study:** 
Based upon the methods described in Klein et al. 2015, we reconstructed the analysis using the subset of protected areas (WDPA 2014) spatially covering a marine area and classified as IUCN I-IV. The WDPA polygons and marine polygons were rasterized to 0.01° and then aggregated to AquaMaps native 0.5° cells, to calculate proportion of protected area and marine within each cell.  After verifying our results using the 2014 AquaMaps dataset, we updated the analysis using the 2015 AquaMaps dataset at a presence threshold of 50% (to compare to Klein et al. directly) and 0% (to better compare with IUCN spatial data).  To analyze MPA coverage against IUCN spatial data, we extracted IUCN polygon weights per 0.5° cell for each species and compared against the protected area raster.  Finally, we combined AquaMaps data (at 0% threshold) and IUCN data, using AquaMaps for the `r n_spp_both` overlapping species and again compared against the protected area raster.

All processing was completed in R (REF), and all code and intermediate data are available on GitHub (REF?) at https://github.com/OHI-Science/IUCN-AquaMaps.
